---
title: "Reproducible Report Example"
author: "Derek H. Ogle"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document
---

```{r setup, include=FALSE}
YEAR <- 2012
knitr::knit_hooks$set(par1 = function(before, options, envir) {
  if (before) par(mar=c(3.5,3.5,1,1),mgp=c(2.1,0.4,0),tcl=-0.2)
})
knitr::opts_chunk$set(echo=FALSE, fig.width=4, fig.height=4, par1=TRUE)
library(FSA)
library(dplyr)
library(captioner)
figs <- captioner(prefix="Figure")
```
```{r getData}
wae <- read.csv("WAE_Escanaba_2011_14.csv") %>%
  select(-Lake,-Assessment) %>%
  rename(len=inches,wt=pounds) %>%
  mutate(sex=mapvalues(sex,from=1:3,to=c("male","female","unknown")),
         sex=factor(sex),
         lcat=lencat(len,w=1),
         loglen=log(len),logwt=log(wt)) %>%
  filterD(year==YEAR)
```

# Introduction
Blah-Blah-Blah

# Findings
```{r}
ttl <- nrow(wae)
sex <- xtabs(~sex,data=wae)
slen <- Summarize(~len,data=wae)
rcum <- rcumsum(xtabs(~lcat,data=wae))
rcump <- rcum/ttl*100
```

A total of `r ttl` Walleye were captured in `r YEAR`, including `r sex[["female"]]` female, `r sex[["male"]]` male, and `r sex[["unknown"]]` unknown sex individuals. The length distribution was right skewed (`r figs("Histogram",display="cite")`) with a mean length of `r formatC(slen[["mean"]],format="f",digits=1)` inches with `r formatC(rcump[["14"]],format="f",digits=1)`% of fish 14 inches or larger and `r formatC(rcump[["20"]],format="f",digits=1)`% of fish 20 inches or larger.

```{r results='hide'}
figs(name="Histogram",paste0("Length frequency histogram for Walleye captured from Escanaba Lake in ",YEAR,"."))
hist(~len,data=wae,w=0.5,xlab="Total Length (inches)",
     ylab="Number of Walleye")
```

`r figs("Histogram")`

